---
title: "7monthcomparison"
author: "Paul Bradshaw"
date: "20/12/2018"
output: html_document
---


# Generating 7-period totals to compare

The data covers the years 2011-12 onwards, split into 13 four-weekly periods.

The latest data covers the first 7 periods of 2018/19.

There are no annual totals, so we need to create those in order to look at change across time.

## Import the data

The data is in a spreadsheet [from the Office of Rail and Road](http://orr.gov.uk/statistics/published-stats/additional-datasets): *Delay minutes, PPM failures and CaSL failures by TOC and Cause, 2011-12 to 2018-19* ([final table](http://orr.gov.uk/__data/assets/excel_doc/0006/25179/delays-by-cause-by-toc.xlsx)) which has been downloaded to the same folder as this notebook. 

We import the first sheet of data (showing minutes of delays by each cause and operator) using the `readxl` package function `read_excel`, removing the initial 12 rows of headings and notes along the way:

```{r}
delayminutes <- readxl::read_excel("delays-by-cause-by-toc20dec2018.xlsx", sheet=1, skip=12, na="")
colnames(delayminutes)
```

The first period starts in the fourth column, D.

We need to find a way to say 'add together all 7 column periods.

```{r}
#Generate a sequence of 'starting point' indexes that relate to each year
#This starts at 4, ends at the index which is the same as the number of columns in the data, and goes in steps of 13
seq(4,length(delayminutes),13)
#Store it
p1indexes <- seq(4,length(delayminutes),13)
#Remove the last item because the last few cols are totals
p1indexes <- head(p1indexes, -1)
#Show the headings for those indexes to test it's worked
colnames(delayminutes)[p1indexes]
#Loop through the indexes, and for each loop through numbers 1 to 12
#Add the resulting column to a running total
for(i in p1indexes){
  #Grab the year from the column name, which is the first 7 characters
  currentyear <- substr(colnames(delayminutes)[i],1,7)
  #Store the minutes for the first column in the range
  yrtotalsofar <- delayminutes[i]
  #Now loop through 1-6 and add each of those to the current index to grab a different column
  for (a in seq(1,6)){
    print(i+a)
    #Add the minutes from that column to the running total
    yrtotalsofar <- yrtotalsofar+delayminutes[i+a]  
  }
  #Once out of that loop, show the last column it grabbed
  print(colnames(delayminutes[i+a]))
  #Create a new column containing the results
  #see https://stackoverflow.com/questions/25165197/r-create-new-column-with-name-coming-from-variable
  delayminutes[,currentyear] <- yrtotalsofar
  #This repeats for each of the 13-period ranges
}
colnames(delayminutes)
```

Export to check

```{r}
write.csv(delayminutes,"delaymins7.csv")
```


That makes the data 105 columns wide, so we'll create a new one with just the years

```{r}
#Calculate the index of the first column of new data
newdataindex <- length(delayminutes)-length(p1indexes)+1
#Test that we can use this to show the last columns of new data, from that position to the last column
delayminutes[newdataindex:length(delayminutes)]
#Bind that to the first 3 columns to create some new data:
delayminutes.justyrs7 <- cbind(delayminutes[1:3],delayminutes[newdataindex:length(delayminutes)])
```

And export:

```{r}
write.csv(delayminutes.justyrs7, "delayminsyrs7.csv")
```