---
title: "Analysing train delay data"
output: html_notebook
---

# Analysing train delay data

This notebook details the process of analysing data on train delays. We are particularly interested in the following questions:

* How many minutes of delay were caused by each identified cause (e.g. tracks, crew etc)?
* How have delays changed over time, in total
* ...and by cause
* ...and by operator
* How has the makeup of each operator's delay types changed over time - are some operators now more likely to have delays because of factors out of their control, or more likely to have delays that they could have directly prevented?

## Import the data

The data is in a spreadsheet [from the Office of Rail and Road](http://orr.gov.uk/statistics/published-stats/additional-datasets): *Delay minutes, PPM failures and CaSL failures by TOC and Cause, 2011-12 to 2018-19* ([final table](http://orr.gov.uk/__data/assets/excel_doc/0006/25179/delays-by-cause-by-toc.xlsx)) which has been downloaded to the same folder as this notebook. 

We import the first sheet of data (showing minutes of delays by each cause and operator) using the `readxl` package function `read_excel`, removing the initial 12 rows of headings and notes along the way:

```{r}
delayminutes <- readxl::read_excel("delays-by-cause-by-toc (6).xlsx", sheet=1, skip=12, na="")
colnames(delayminutes)
```

There are a lot of columns here - 98 in all.

```{r}
summary(delayminutes)
```


## Generating annual totals

The data covers the years 2011-12 onwards, split into 13 four-weekly periods.

There are no annual totals, so we need to create those in order to look at change across time.

The first period starts in the fourth column, D.

We could add those together manually...

```{r}
#Show the first period
delayminutes[4]
#Show the second period
delayminutes[5]
#Add the two together
delayminutes[4]+delayminutes[5]
```

...but that's going to be laborious. We need to find a way to say 'add together all 13 columns.

```{r}
#Generate a sequence of 'starting point' indexes that relate to each year
#This starts at 4, ends at the index which is the same as the number of columns in the data, and goes in steps of 13
seq(4,length(delayminutes),13)
#Store it
p1indexes <- seq(4,length(delayminutes),13)
#Remove the last item because the last few cols are totals
p1indexes <- head(p1indexes, -1)
#Show the headings for those indexes to test it's worked
colnames(delayminutes)[p1indexes]
#Loop through the indexes, and for each loop through numbers 1 to 12
#Add the resulting column to a running total
for(i in p1indexes){
  #Grab the year from the column name, which is the first 7 characters
  currentyear <- substr(colnames(delayminutes)[i],1,7)
  #Store the minutes for the first column in the range
  yrtotalsofar <- delayminutes[i]
  #Now loop through 1-12 and add each of those to the current index to grab a different column
  for (a in seq(1,12)){
    print(i+a)
    #Add the minutes from that column to the running total
    yrtotalsofar <- yrtotalsofar+delayminutes[i+a]  
  }
  #Once out of that loop, show the last column it grabbed
  print(colnames(delayminutes[i+a]))
  #Create a new column containing the results
  #see https://stackoverflow.com/questions/25165197/r-create-new-column-with-name-coming-from-variable
  delayminutes[,currentyear] <- yrtotalsofar
  #This repeats for each of the 13-period ranges
}
colnames(delayminutes)
```

Export to check

```{r}
write.csv(delayminutes,"delaymins.csv")
```


That makes the data 105 columns wide, so we'll create a new one with just the years

```{r}
#Calculate the index of the first column of new data
newdataindex <- length(delayminutes)-length(p1indexes)+1
#Test that we can use this to show the last columns of new data, from that position to the last column
delayminutes[newdataindex:length(delayminutes)]
#Bind that to the first 3 columns to create some new data:
delayminutes.justyrs <- cbind(delayminutes[1:3],delayminutes[newdataindex:length(delayminutes)])
```

And export:

```{r}
write.csv(delayminutes.justyrs, "delayminsyrs.csv")
```

## Query the data 

What questions do we want to ask?

* How many minutes of delays did traincrew-related causes responsible for?
* How many were due to severe weather, etc.?
* How many were due to 'fleet'?
* How many were 'TOC-on-self' ("Issues which the train company could have prevented")
* What were these figures for each operator (TOC)?
* How have these figures changed for each operator (TOC)?
* How have the proportions changed by cause or by TOC between 2011 and the latest figures?

Let's get the data frame working in `sqldf`:

```{r}
#SQL interprets periods to indicate a table.field relationship, so we need to create a data frame without a period
#We need to remove the grand totals anyway so let's do that too:
delayminutesyrs <- subset(delayminutes.justyrs, delayminutes.justyrs$`Delay Type` != "All Delay Minutes")
#Show a few rows first
sqldf::sqldf('select * from delayminutesyrs limit 10')
```

Now let's create a pivot table

```{r}
sqldf::sqldf('SELECT `Train Operating Company (TOC)`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
             GROUP BY `Train Operating Company (TOC)`
             ORDER BY pcchange1118 DESC')
totalminsbytoc <- sqldf::sqldf('SELECT `Train Operating Company (TOC)`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
             GROUP BY `Train Operating Company (TOC)`
             ORDER BY pcchange1118 DESC')
```

We can make this easier by generating the query:

```{r}
#Create an empty string to add to
querystring <- ""
#Loop through the list of column names from col 4 on, and add them to the string
for (i in colnames(delayminutesyrs)[4:length(delayminutesyrs)]){
  print(i)
  #Insert characters needed for SQL and to SUM
  querystring <- paste(querystring,i,sep="`), SUM(`")
}
querystring
#Get rid of the extra chars at the start
querystring <- substr(querystring,5,nchar(querystring))
querystring
#Add chars at the end
querystring <- paste(querystring,"`)",sep="")
querystring
```

Now to use it:

```{r}
#Paste the string along with the rest of a query
sqldf::sqldf(paste("SELECT `Train Operating Company (TOC)`,", querystring, "FROM delayminutesyrs 
             GROUP BY `Train Operating Company (TOC)`"))

#Store in a data frame
yeartotals <- sqldf::sqldf(paste("SELECT `Train Operating Company (TOC)`,",querystring,"FROM delayminutesyrs 
             GROUP BY `Train Operating Company (TOC)`"))
```


### Looking at train crew only

```{r}
sqldf::sqldf('
  SELECT *, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Delay Type` IS "TOC-on-Self" 
  AND `Category of Delay` IS "Traincrew"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
traincrewselfonly <- sqldf::sqldf('
  SELECT *, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Delay Type` IS "TOC-on-Self" 
  AND `Category of Delay` IS "Traincrew"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
write.csv(traincrewselfonly,"traincrewselfonly.csv")
```

And the totals for both types of traincrew delay (TOC on self and TOC on TOC)

```{r}
#delayminutesyrs$
sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Category of Delay`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Category of Delay` IS "Traincrew"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
traincrewall <- sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Category of Delay`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Category of Delay` IS "Traincrew"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
write.csv(traincrewall,"traincrewall.csv")
```


### Looking at weather

Let's fine out what the heading is. There should be 23 results for those which are TOC-on-self only, and 46 where it is used in another delay type too (TOC-on-TOC):

```{r}
table(delayminutesyrs$`Category of Delay`)
```

Weather is only used for one delay type so we don't need to specify that.

```{r}
sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Category of Delay`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Category of Delay` IS "Severe Weather, Autumn,  & Structures"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
weather <- sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Category of Delay`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Category of Delay` IS "Severe Weather, Autumn,  & Structures"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
write.csv(weather,"weather.csv")
```

### Delays caused by tracks

The same applies to track:

```{r}
sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Category of Delay`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Category of Delay` IS "Track"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
trackdelays <- sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Category of Delay`, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Category of Delay` IS "Track"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
write.csv(trackdelays,"trackdelays.csv")
```
### Delays caused by fleet

Fleet delays can be either TOC-on-self or TOC-on-TOC.

```{r}
sqldf::sqldf('
  SELECT *, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Delay Type` IS "TOC-on-Self" 
  AND `Category of Delay` IS "Fleet"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
fleetdelays.self <- sqldf::sqldf('
  SELECT *, SUM(`2011-12`), SUM(`2017-18`), SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118 FROM delayminutesyrs 
  WHERE `Delay Type` IS "TOC-on-Self" 
  AND `Category of Delay` IS "Fleet"
  GROUP BY `Train Operating Company (TOC)`
  ORDER BY pcchange1118 DESC
  ')
```

### TOC-on-Self as a % of total - combining tables

Now we switch perspective a little. What is the total of all the TOC-on-Self delay minutes?

First we grab the totals of each TOC which we need to calculate a % of each total. We have already created a very similar data frame called `totalminsbytoc`, but it's worth doing again here (and that has 5 columns, this will only have 3).

```{r}
#Calculate the totals of each TOC. We rename the columns to avoid confusion when they are later merged
totalsbytoc <- sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`, SUM(`2011-12`) AS toctotal1112, SUM(`2017-18`) AS toctotal1718
  FROM delayminutesyrs 
  GROUP BY `Train Operating Company (TOC)` 
  ')
totalsbytoc
```

Next we generate a table showing the totals by delay type and TOC:

```{r}
#give totals name to distinguish again
sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Delay Type`, SUM(`2011-12`) AS typetotal1112, SUM(`2017-18`) AS typetotal1718, SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118
  FROM delayminutesyrs 
  GROUP BY `Train Operating Company (TOC)`, `Delay Type`
  ORDER BY `Train Operating Company (TOC)`, pcchange1118 DESC
  ')
typebreakdown <- sqldf::sqldf('
  SELECT `Train Operating Company (TOC)`,`Delay Type`, SUM(`2011-12`) AS typetotal1112, SUM(`2017-18`) AS typetotal1718, SUM(`2017-18`)-SUM(`2011-12`) AS change1118, (SUM(`2017-18`)-SUM(`2011-12`))/SUM(`2011-12`) AS pcchange1118
  FROM delayminutesyrs 
  GROUP BY `Train Operating Company (TOC)`, `Delay Type`
  ORDER BY `Train Operating Company (TOC)`, pcchange1118 DESC
  ')
write.csv(typebreakdown,"typebreakdown.csv")
```
Then we join that table and the one with the totals by TOC

```{r}
#Combine both tables, creating some new fields which calculate the % of each TOC's total delays attributable to each type in each period.
sqldf::sqldf('
  SELECT *, (typetotal1112 / toctotal1112) as pcofTOCtotal1112, (typetotal1718 / toctotal1718) as pcofTOCtotal1718, (typetotal1718 / toctotal1718)-(typetotal1112 / toctotal1112) as pointschange
  FROM typebreakdown 
  LEFT JOIN totalsbytoc
  ON typebreakdown.`Train Operating Company (TOC)` = totalsbytoc.`Train Operating Company (TOC)`
  ')
#Now store the results
toctypepcchange <- sqldf::sqldf('
  SELECT *, (typetotal1112 / toctotal1112) as pcofTOCtotal1112, (typetotal1718 / toctotal1718) as pcofTOCtotal1718, (typetotal1718 / toctotal1718)-(typetotal1112 / toctotal1112) as pointschange
  FROM typebreakdown 
  LEFT JOIN totalsbytoc
  ON typebreakdown.`Train Operating Company (TOC)` = totalsbytoc.`Train Operating Company (TOC)`
  ')
#Export as CSV
write.csv(toctypepcchange, "toctypepcchange.csv")
```

We can see which operators have had the biggest increase in TOC-on-self delays by filtering and sorting

```{r}
#Create a subset that only shows the TOC on self delay hours
toctypepcchange.tosonly <- subset(toctypepcchange, toctypepcchange$`Delay Type` == "TOC-on-Self")
#Attach it to sort it
attach(toctypepcchange.tosonly)
#Sort by pointschange, descending
toctypepcchange.ordered <- toctypepcchange.tosonly[order(-pointschange),] 
#Show the results
toctypepcchange.ordered
#Detach it
detach(toctypepcchange.tosonly)
```


